The Functional Therapeutic Chemical Classification System

### 2013-08-15 document.tex
supp figures all included
Signed-off-by: loopasam <samuel.croset@gmail.com>
references added
Signed-off-by: loopasam <samuel.croset@gmail.com>

\title[short Title]{The Functional Therapeutic Chemical Classification System}
\author[Croset \textit{et~al}]{Samuel Croset\,$^{1, \footnote{to whom correspondence should be addressed}}$, John Overington\,$^{1}$ and Dietrich Rebholz-Schuhmann\,$^{1}$
}
\address{$^{1}$European Molecular Biology Laboratory
European Bioinformatics Institute (EMBL-EBI), Wellcome Trust Genome Campus, Hinxton, Cambridge CB10 1SD,
United Kingdom\\}

@@

Drug repurposing is the application of known drugs and compounds to new indications (i.e., new diseases) (cite wikipedia and better source). 

@@

interesting repurposing opportunities (http://www.ncbi.nlm.nih.gov/pubmed/23340113). For instance, sildenafil was initially developed 
against angina and has been repurposed towards erectile dysfunction during the clinical trials (cite). Approved compounds are 

@@

There is an increasing number of approaches to predict repurposing opportunities using computational methods (cite). Most 
methods operate on the profiles of physicochemical descriptors derived from molecular structures (cite). Other methods characterise 
the drugs on more abstract levels, such as the gene expression signature (cite) or via the reported side-effects (cite). These approaches 

@@

A feature of particular interest to describe drugs is the mode of action (MoA). The MoA describes a functional or anatomical change, 
at the cellular level, resulting from the exposure of a living organism to a substance (cite wikipedia or better source). For instance 

@@

Nonetheless, the meaning of a concept can be formalised with controlled vocabularies and ontologies (cite); originating from description logics, such frameworks help to define the mathematical meaning of symbols and strings of characters. In an ontology or knowledge base, concepts are organised and linked following the logical type of relation they have between them. In the Gene Ontology (GO) for example (cite), biological processes and molecular functions are manually curated in this resource and the meaning of terms is specified by the relations linking two GO terms.
MoA definitions are present in other classifications such as the Medical Subject Headings (cite) or the Chemical Entities of 
Biological Interest (cite) for example. The Anatomical Therapeutic Chemical Classification System (ATC) also describes to some 

@@

of various biomedical databases using description logics and automated reasoning. Over 20’000 new MoA categories are defined 

@@

 Text Text Text Text. Figure~\ref{fig:01} shows that the above method  Text 
   Text Text.  \citep{Bag01} wants to know about text follows

@@

The knowledge base behind the FTC is built by integrating information coming from various sources. 
The GO terms serve as template to create the FTC categories describing the MoAs; DrugBank provides the 
known links between drugs and their protein targets. Uniprot maps targets to their respective GO annotations. 
Drugs are further assigned into mode of action categories based of the OWL constructs and axioms defined in the 
FTC (see Availability and Implementation). A reasoner, a program capable of understanding such axioms, performed this task. 
The process to build the FTC is summarised in Figure 0.

@@

It takes around 5 seconds on a modern 
laptop (4 cores processor with 4 Gb of memory) to classify the knowledge base using the ELK reasoner (cite). 
Other OWL reasoners have been tried (e.g, Hermit (cite), Pellet (cite), etc...) but none of them were suitable for this purpose; 
the classification time was always superior to a few minutes (data not shown). The FTC has a taxonomic structure as illustrated 
on Figure 1, which arises when the reasoner classifies the knowledge base. Categories can have multiple parents and multiple children. 
The reader can browse and access the content of the FTC online at {{https://www.ebi.ac.uk/chembl/ftc}}.

@@

The subgroups B, C and D are inhibitors of the cholinergic system and some of them, such as galantamine (DB00674), are already 
investigated to treat Alzheimer’s disease and other related dementias (pubchem). This class of agent is in line with the 
cholinergic hypothesis (cite PMID:10071091), stating that Alzheimer’s disease could be caused by dysfunctions in the processing of acetylcholine. The subgroup E is exclusively composed of barbiturates (central nervous system depressants). The presence of this pharmacological class of compounds as an Alzheimer treatment is more surprising, as very little literature reports on it. Further investigations reveal that the neuronal acetylcholine receptor subunit alpha-7 (P36544), a common off-target of barbiturates, binds beta-amyloids with high affinity. As beta-amyloids are themselves strongly involved in the pathology (cite http://europepmc.org/abstract/MED/10681545), barbiturates could affect the state of the condition, similarly to cholinergic inhibitors.
The group F contains four compounds; nicotine and varenicline have been grouped because of similar pharmacology (group G). 
Nicotine has been shown to improve some of the symptoms of Alzheimer’s disease (PMID:1410164), it is therefore expected to 
find this molecule in the predictions. Varenicline possess a pharmacology related to that of nicotine, which would explain 
the presence of the drug in this category. The two remaining drugs of group F are respectively pralidoxime and dipivefrin; 
little information is available regarding their potential action against the disease, yet these compounds seem linked to the 
cholinergic hypothesis and could be considered for experimental testing.
The groups H and I contain respectively one molecule each. These compounds have been classified as agents perturbing some of 
the physiological function of the Tau protein, key actor in Alzheimer's disease (cite). Vorinostat (group H) is currently 
indicated for the treatment of cutaneous manifestations in patients with T-cell lymphoma (drugbank), yet a study has 
shown in vivo (mouse model) the potential of the drug and other histone deacetylase inhibitors in regards to memory 
deficit (cite http://www.nature.com/npp/journal/v35/n4/full/npp2009197a.html). The presence of lithium (group I) was 
unexpected and suspected of being a false positive at first glance. However a recent study demonstrated a long-term 
protective effect for the ion in regards to Alzheimer’s disease (PMID:22500970 and 21525519). The last group (J) contains 
ezetimibe and hesperetin. These two compounds are primarily used as cholesterol lowering agents (statins). As cholesterol 
metabolism in the brain appears to be related to dementia, statins are believed to prevent or improve the symptoms of the 
patients. Although early studies (PMID:17877925) have failed to clearly show a beneficial effect, the investigation is still open.
From the examples briefly presented above, reported and confirmed by the literature, the FTC appears to be suitable to 
identify real repurposing hypotheses tailored to a disease. The MoAs help to retrieve the compounds which might impact 
the treatment of a given set of symptoms.

@@

%\bibliographystyle{natbib}
%\bibliographystyle{achemnat}
%\bibliographystyle{plainnat}
%\bibliographystyle{abbrv}
%\bibliographystyle{bioinformatics}
%
%\bibliographystyle{plain}
%
%\bibliography{Document}

\begin{thebibliography}{}
\bibitem[Bofelli {\it et~al}., 2000]{Boffelli03} Bofelli,F., Name2, Name3 (2003) Article title, {\it Journal Name}, {\bf 199}, 133-154.

\bibitem[Bag {\it et~al}., 2001]{Bag01} Bag,M., Name2, Name3 (2001) Article title, {\it Journal Name}, {\bf 99}, 33-54.

\bibitem[Yoo \textit{et~al}., 2003]{Yoo03}
Yoo,M.S. \textit{et~al}. (2003) Oxidative stress regulated genes
in nigral dopaminergic neurnol cell: correlation with the known
pathology in Parkinson's disease. \textit{Brain Res. Mol. Brain
Res.}, \textbf{110}(Suppl. 1), 76--84.

\bibitem[Lehmann, 1986]{Leh86}
Lehmann,E.L. (1986) Chapter title. \textit{Book Title}. Vol.~1, 2nd edn. Springer-Verlag, New York.

\bibitem[Crenshaw and Jones, 2003]{Cre03}
Crenshaw, B.,III, and Jones, W.B.,Jr (2003) The future of clinical
cancer management: one tumor, one chip. \textit{Bioinformatics},
doi:10.1093/bioinformatics/btn000.

\bibitem[Auhtor \textit{et~al}. (2000)]{Aut00}
Auhtor,A.B. \textit{et~al}. (2000) Chapter title. In Smith, A.C.
(ed.), \textit{Book Title}, 2nd edn. Publisher, Location, Vol. 1, pp.
???--???.

\bibitem[Bardet, 1920]{Bar20}
Bardet, G. (1920) Sur un syndrome d'obesite infantile avec
polydactylie et retinite pigmentaire (contribution a l'etude des
formes cliniques de l'obesite hypophysaire). PhD Thesis, name of
institution, Paris, France.

\end{thebibliography}
\end{document}

>>>>

\title[The Functional Therapeutic Chemical Classification System]{The Functional Therapeutic Chemical Classification System}
\author[Croset \textit{et~al}]{Samuel Croset\,$^{1, \footnote{to whom correspondence should be addressed}}$, John Overington\,$^{1}$ and Dietrich Rebholz-Schuhmann\,$^{1}$
}
\address{$^{1}$European Molecular Biology Laboratory
European Bioinformatics Institute (EMBL-EBI), Wellcome Trust Genome Campus, Hinxton, Cambridge CB10 1SD,
United Kingdom\\}

@@

Drug repurposing is the use of known active compounds for new therapeutic indications \citep{Sanseau01072011}. 

@@

interesting repurposing opportunities \citep{Medina-Franco2013}. For instance, sildenafil was initially developed 
against angina and has been repurposed towards erectile dysfunction during the clinical trials \citep{Ashburn2004}. Approved compounds are 

@@

There is an increasing number of approaches to predict repurposing opportunities using computational methods \citep{Sanseau01072011}. Most 
methods operate on the profiles of physicochemical descriptors derived from molecular structures \citep{Haupt2011}. Other methods characterise 
the drugs on more abstract levels, such as the gene expression signature \citep{Iorio2010} or 
via the reported side-effects \citep{Campillos2008}. 
These approaches 

@@

A feature of particular interest to describe drugs is the mode of action (MoA). According to Wikipedia, 
the MoA describes a functional or anatomical change, 
at the cellular level, resulting from the exposure of a living organism to a substance. For instance 

@@

Nonetheless, the meaning of a concept can be formalised with controlled vocabularies and ontologies \citep{Gruber1995}; 
originating from description logics, such frameworks help to define the mathematical meaning of symbols and strings of characters. 
In an ontology or knowledge base, concepts are organised and linked following the logical type of relation they have between them. 
In the Gene Ontology (GO) for example \citep{Ashburner2000}, biological processes and molecular functions are manually curated in this resource 
and the meaning of terms is specified by the relations linking two GO terms.
MoA definitions are present in other classifications such as the Medical Subject Headings \citep{Nelson2004} or the Chemical Entities of 
Biological Interest \citep{Hastings2012} for example. The Anatomical Therapeutic 
Chemical Classification System (ATC) \citep{world2000anatomical} also describes to some 

@@

of various biomedical databases using description logics and automated reasoning. Over 20'000 new MoA categories are defined 

@@

@@

The knowledge base behind the FTC is built by integrating information coming from various sources. 
The GO terms serve as template to create the FTC categories describing the MoAs; DrugBank provides the 
known links between drugs and their protein targets. Uniprot maps targets to their respective GO annotations. 
Drugs are further assigned into mode of action categories based of the OWL constructs and axioms defined in the 
FTC (see Availability and Implementation). A reasoner, a program capable of understanding such axioms, performed this task. 
The process to build the FTC is summarised in Figure~\ref{fig:01}.

@@

It takes around 5 seconds on a modern 
laptop (4 cores processor with 4 Gb of memory) to classify the knowledge base using the ELK reasoner (cite). 
Other OWL reasoners have been tried (e.g, Hermit (cite), Pellet (cite), etc...) but none of them were suitable for this purpose; 
the classification time was always superior to a few minutes (data not shown). The FTC has a taxonomic structure as illustrated 
on Figure~\ref{fig:02}, which arises when the reasoner classifies the knowledge base. Categories can have multiple parents and multiple children. 
The reader can browse and access the content of the FTC online at {{https://www.ebi.ac.uk/chembl/ftc}}.

@@

The subgroups B, C and D are inhibitors of the cholinergic system and some of them, such as galantamine (DB00674), are already 
investigated to treat Alzheimer’s disease and other related dementias. This class of agent is in line with the 
cholinergic hypothesis \citep{Francis1999}, stating that Alzheimer’s disease could be caused by dysfunctions in the processing of 
acetylcholine. The subgroup E is exclusively composed of barbiturates (central nervous system depressants). 
The presence of this pharmacological class of compounds as an Alzheimer treatment is more surprising, as very little 
literature reports on it. Further investigations reveal that the neuronal acetylcholine receptor subunit alpha-7, a 
common off-target of barbiturates, binds beta-amyloids with high affinity \citep{Wang2000}. As beta-amyloids are themselves strongly involved 
in the pathology, barbiturates could affect the state of the condition \citep{Wang2000}, similarly to cholinergic inhibitors.
The group F contains four compounds; nicotine and varenicline have been grouped because of similar pharmacology (group G). 
Nicotine has been shown to improve some of the symptoms of Alzheimer’s disease \citep{Jones1992}, it is therefore expected to 
find this molecule in the predictions. Varenicline possess a pharmacology related to that of nicotine, which would explain 
the presence of the drug in this category. The two remaining drugs of group F are respectively pralidoxime and dipivefrin; 
little information is available regarding their potential action against the disease, yet these compounds seem linked to the 
cholinergic hypothesis and could be considered for experimental testing.
The groups H and I contain respectively one molecule each. These compounds have been classified as agents perturbing some of 
the physiological function of the Tau protein, key actor in Alzheimer's disease \citep{Grundke-Iqbal1986}. Vorinostat (group H) is currently 
indicated for the treatment of cutaneous manifestations in patients with T-cell lymphoma, yet a study has 
shown in vivo (mouse model) the potential of the drug and other histone deacetylase inhibitors in regards to memory 
deficit \citep{Kilgore2010}. The presence of lithium (group I) was 
unexpected and suspected of being a false positive at first glance. However a recent study demonstrated a long-term 
protective effect for the ion in regards to Alzheimer’s disease \citep{Young2011}. The last group (J) contains 
ezetimibe and hesperetin. These two compounds are primarily used as cholesterol lowering agents (statins). As cholesterol 
metabolism in the brain appears to be related to dementia, statins are believed to prevent or improve the symptoms of the 
patients. Although early studies \citep{Wolozin2004} have failed to clearly show a beneficial effect, the investigation is still open.
From the examples briefly presented above, reported and confirmed by the literature, the FTC appears to be suitable to 
identify real repurposing hypotheses tailored to a disease. The MoAs help to retrieve the compounds which might impact 
the treatment of a given set of symptoms.

@@

\bibliographystyle{natbib}
\bibliographystyle{achemnat}
\bibliographystyle{plainnat}
\bibliographystyle{abbrv}
\bibliographystyle{bioinformatics}

\bibliographystyle{plain}

\bibliography{document}

%\begin{thebibliography}{}
%\bibitem[Bardet, 1920]{Bar20}
%Bardet, G. (1920) Sur un syndrome d'obesite infantile avec
%polydactylie et retinite pigmentaire (contribution a l'etude des
%\end{thebibliography}
\end{document}
